---
title: "A fast algorithm for tau-estimators"
author: "Matias Salibian"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A fast algorithm for tau-estimators for linear regression 

This repository contains `R` (in file FastTau.R) and `MATLAB` / `OCTAVE` 
(in file FastTau.m) code implementing
a fast algorithm to compute tau-regression estimators
as proposed in 
[Salibian-Barrera, M., Willems, G. and 
Zamar, R.H. (2008)](http://dx.doi.org/10.1198/106186008X343785). 

Below is an example illustrating its use in `R`. We first 
read the function:
```{r initial}
source('FastTau.R')
```
We will apply it to the well-known Boston Housing data set. 
```{r datasetup}
data(Boston, package='MASS')
x <- model.matrix(medv ~ ., data=Boston)
y <- Boston$medv
tauest <- FastTau(x=x, y=y, N=500, kk=2, tt=5, rr=2, approximate=0, seed=456)
```
The arguments of `fast.s` are:

* y: response vector (n x 1)
* x: covariates matrix (n x p), possibly including a column of ones (intercept term)
* N: number of elemental sub-set starts 
* kk: number of initial IRWLS steps to be applied to each elemental candidate
* tt: number of best partially iterted candidates to further RWLS-iterate until convergence
* approximate: if 0, fully compute S-scale when evaluating the objective function, otherwise use an approximation 
* rr: if `approximate = 1` above, rr is number of iterations used for the scale approximation
* seed: seed used in the generation of the random sub-samples

This code computes a tau-estimate tuned to have 95% efficiency and 50% breakdown-point, 
using a rho-function that produces an estimate with small asymptotic maximum bias. 
The efficiency and breakdown point are controlled with the
variables 
`c1`,
`b1`,
`c2` and 
`b2` (lines 35-38 in the file FastTau.R). 
See [Salibian-Barrera, M., Willems, G. and 
Zamar, R.H. (2008)](http://dx.doi.org/10.1198/106186008X343785). 
for more details. 


We can compare the S- and LS-estimators:
```{r compare}
lmest <- lm(medv ~ ., data=Boston)
round(cbind(tauest$beta, coef(lmest)), 3)
```
The estimated residual scales are
```{r compareses}
round(c(tauest$scale, summary(lmest)$sigma), 2)
```

I will later add here an illustration of the better predictions obtained
with the robust estimator.

